### 2010-11-16 paper.tex
original version

\documentclass[12pt,letterpaper, singlecolumn]{article}
\usepackage{latex8}
\usepackage{times}
\usepackage{cite}
\usepackage{subfigure}
\usepackage{psfig}
\usepackage[pdftex]{graphicx}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{comment}
\usepackage[small,compact]{titlesec}
\usepackage[small,it]{caption}
\usepackage{wrapfig}

\begin{document}

\title{CSE530 Project Mid-Point Progress Report\\
Bandwidth-Aware Memory Hierarchy Design with Hybrid Memory Technologies}

\author{Cong Xu and Jishen Zhao\\
\{czx102, juz138\}@psu.edu\vspace{-10pt}
}
\maketitle

\begin{large}

% ------------------------------------------------------------------------------

\section{Project Description}

One of the challenges for modern chip-multiprocessor~(CMP) design is the growing bandwidth gap
between processor cores and off-chip main memory. Emerging memory technologies
bring in potential opportunities to fill this gap. The memory hierarchy,
however, need to be carefully designed in order to bridge the this gap with low
performance and power overhead.

In this project, we would like to explore the optimal memory hierarchy design
with various memory technologies specifically for bandwidth-bounded
applications. We will first estimate the performance of an application on a
baseline system, in terms of the bandwidth demanded by the application for a
range of memory capacities and the bandwidth that can be provided by the system
with such memory capacities. We would like to find out whether the bandwidth
provided by the memory hierarchy can satisfy the demand. If it is not satisfied,
we can modify the memory hierarchy, for example, by changing the capacity of
existing cache and memories, or by adding an extra level of cache or memory.
Based on our exploration, we would like to find a memory hierarchy design
solution optimized for bandwidth, including the number of cache and memory
levels, as well as the memory technology, the bandwidth, and the capacity used
for each level. In addition, we will enhance the memory hierarchy design from
the energy-efficiency point of view by constraining the total power consumption
of the memory system.\vspace{0.15in}

% ------------------------------------------------------------------------------

\section{Motivation}

Many of modern chip-multiprocessors are designed to perform well on various
applications to achieve high performance by exploiting their inherent
parallelism. Such systems support large number of threads and single instruction
multiple data~(SIMD) execution, which puts a lot of pressure on the memory
system. Memory latency is typically not a bottleneck, since the latency can be
hidden via multi-threading or hardware prefetching. To this end, bandwidth becomes a
potential bottleneck. A high rate of computing often brings in a high rate of
data transitions. In some cases, the working set of an application fits in the
on-die caches, which can typically provide sufficient bandwidth to keep up with
the processing cores. However, if the working set does not fit in the on-die
caches, the main memory needs to provide much of the data. Since the bandwidth
of off-chip main memory is quite limited, applications with such working sets
are potentially bandwidth-bound. Therefore, it is crucial to design the memory
hierarchy to overcome the bandwidth limitation.

It is known that performance improvement of a computing system can be achieved
via multiple memory levels. Adding an extra level of memory, however, can also
help alleviate the bandwidth bottleneck of off-chip memory. In addition,
emerging memory technologies such as Magnetoresistive Random Access
Memory~(MRAM), Phase-Change Memory~(PCM), Resistive Random Access Memory~(RRAM),
etc., have shown potential to to be used as on-chip caches and the main
memory~\cite{Sun:2009:MRAM-L2-CMP}. Therefore, we would like to explore how to
enhance the memory hierarchy from the bandwidth point of view. Specifically, we
would like to examine (1) the number of levels in the optimal memory hierarchy,
(2) the appropriate memory technology of each level, and (3) the capacity and
bandwidth of each level. We will also explore the energy-efficiency of the
memory hierarchy design constrain, and the memory hierarchy design within a
fixed power budge.\vspace{0.15in}

% -----------------------------------------------------------------------------

\section{Experimental Setup}

We will use our modified CACTI~\cite{CACTI} to estimate the bandwidths provided
by different memory technologies. We will also use
Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
design method. The benchmarks will be selected from bandwidth-bound
applications, which possibly include benchmarks from SPEC
CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}

\section{Experiments}

In this section, we describe the experimental results we have obtained so far
using a modified version of CACTI~\cite{CACTI}. \vspace{-0.5in}

\begin{figure*}[htbp]
% The "!" means to maintain the aspect ratio.   
\centering
\includegraphics[width=3in]{figures/read-bw}
\includegraphics[width=3in]{figures/write-bw}\\

\vspace*{-1in}
\hspace{0.02in}
\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}

\caption{Read and write bandwidths provided by different memory technologies.
(a) Read bandwidth provided by different memory technologes. (b) Write bandwidth
provided by different memory technologies.}
\label{fig:memory-bw}
\end{figure*}

\begin{figure*}[htbp]
% The "!" means to maintain the aspect ratio.   
\centering
\includegraphics[width=2in]{figures/read-energy}
\includegraphics[width=2in]{figures/write-energy}
\includegraphics[width=2in]{figures/total-energy}\\

\vspace*{-0.5in}
\hspace{0.02in}
\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}
\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (c)}}

\caption{Dynamic energy consumption with the provided bandwidths of different
  memory technologies. (a) Read energy of different memory technologes. (b)
  Write energy of different memory technologies. (c) Total energy consumption
  (with 10\% write) of different memory technologies.}
\label{fig:memory-energy}
\end{figure*}

First of all, we estimate the read and write bandwidths that can be provided by
different memory technologies. Figure~\ref{fig:memory-bw} shows the results,
with both x- and y-values in \emph{log} scale. The figure illustrates both the
provided read and write bandwidth as a function of memory capacity. Each of the
memory technologies actually provide nearly the same read bandwidths, as is
shown in figure~\ref{fig:memory-bw}(a). On the other hand, a straight forward
observation from figure~\ref{fig:memory-bw}(b) is that the write bandwidth
varies among different memory technologies. The shape of the SRAM write
bandwidth curve is very similar to the read bandwidth curve. The write bandwidth
curves of the other three memory technologies appear to be very different. The
reason is that write latencies of the three non-volatile memories are much
higher than read latencies. Another observation from
figure~\ref{fig:memory-bw}(b) is that the curves cross to each other at
different locations. Based on these two observations, we would like to design a
bandwidth-aware hybrid memory hierarchy, which always provides the high memory
bandwidth with the given capacity. For example, we probably would like to use
1) SRAM when the memory capacity is under 256MB, 2) MRAM or RRAM between
256MB and 1GB (if only bandwidth is considered), and 3) PCRAM when the memory
cacity is over 1GB.

In addition, we would like to put an upper bound to the total dynamic energy
consumption of the memory hierarchy. Figure~\ref{fig:memory-energy} shows the
estimation of dynamic power consumptions of different memory technologies with
the provided bandwidths. In figure~\ref{fig:memory-energy}(c), we estimate the total
dynamic energy with 10\% write activities.

% \bibliographystyle{IEEEtran}
\bibliographystyle{latex8}
\bibliography{./bib/ref}

\end{large}

\end{document}

### 2010-11-17 trunk/paper.tex
paper.tex
Re-organize the paper structure and initialize some sections

--- 2010-11-16 paper.tex
@@
-\documentclass[12pt,letterpaper, singlecolumn]{article}
+\documentclass[10pt,letterpaper,singlecolumn]{article}
+\usepackage{setspace}
+\doublespacing
 \usepackage{latex8}
@@
 
-\title{CSE530 Project Mid-Point Progress Report\\
+\title{CSE530 Project Final Report\\
 Bandwidth-Aware Memory Hierarchy Design with Hybrid Memory Technologies}
@@
 
-% ------------------------------------------------------------------------------
+\begin{abstract}
+\boldmath
 
-\section{Project Description}
-
-One of the challenges for modern chip-multiprocessor~(CMP) design is the growing bandwidth gap
-between processor cores and off-chip main memory. Emerging memory technologies
-bring in potential opportunities to fill this gap. The memory hierarchy,
-however, need to be carefully designed in order to bridge the this gap with low
-performance and power overhead.
-
-In this project, we would like to explore the optimal memory hierarchy design
-with various memory technologies specifically for bandwidth-bounded
-applications. We will first estimate the performance of an application on a
-baseline system, in terms of the bandwidth demanded by the application for a
-range of memory capacities and the bandwidth that can be provided by the system
-with such memory capacities. We would like to find out whether the bandwidth
-provided by the memory hierarchy can satisfy the demand. If it is not satisfied,
-we can modify the memory hierarchy, for example, by changing the capacity of
-existing cache and memories, or by adding an extra level of cache or memory.
-Based on our exploration, we would like to find a memory hierarchy design
-solution optimized for bandwidth, including the number of cache and memory
-levels, as well as the memory technology, the bandwidth, and the capacity used
-for each level. In addition, we will enhance the memory hierarchy design from
-the energy-efficiency point of view by constraining the total power consumption
-of the memory system.\vspace{0.15in}
+\end{abstract}
 
@@
 
-\section{Motivation}
+\section{Introduction}
 
@@
 via multiple memory levels. Adding an extra level of memory, however, can also
-help alleviate the bandwidth bottleneck of off-chip memory. In addition,
-emerging memory technologies such as Magnetoresistive Random Access
-Memory~(MRAM), Phase-Change Memory~(PCM), Resistive Random Access Memory~(RRAM),
-etc., have shown potential to to be used as on-chip caches and the main
-memory~\cite{Sun:2009:MRAM-L2-CMP}. Therefore, we would like to explore how to
-enhance the memory hierarchy from the bandwidth point of view. Specifically, we
+help alleviate the bandwidth bottleneck of off-chip memory. Specifically, we
 would like to examine (1) the number of levels in the optimal memory hierarchy,
@@
 % -----------------------------------------------------------------------------
+\section{Motivation}
 
-\section{Experimental Setup}
+Throughput computing involves performing a huge number of calculations with a large amount of parallelism.
+Throughput computing applications span many domains and are already critical on a variety of platforms from high-performance computing (HPC) machines to commercial servers to client machines.
 
-We will use our modified CACTI~\cite{CACTI} to estimate the bandwidths provided
-by different memory technologies. We will also use
-Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
-design method. The benchmarks will be selected from bandwidth-bound
-applications, which possibly include benchmarks from SPEC
-CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}
+Systems designed to perform well on throughput computing applications achieve high performance by exploiting their inherent parallelism.
+These systems support large numbers of threads and/or use wide SIMD execution.
+This puts a lot of pressure on the memory system.
+Fortunately, memory latency is typically not a bottleneck since the latency can be hidden via multithreading or hardware prefetching, since the data access patterns are relatively simple.
+On the other hand, bandwidth is a potential bottleneck.
 
-\section{Experiments}
+Many throughput computing applications have inherently large working sets (i.e., tens to hundreds of MB).
+These are unlikely to fit in conventional on-die SRAM caches for the foreseeable future.
+Further, unlike more traditional workloads (e.g., those similar to SPEC or TPC benchmarks), many throughput computing applications see a sharp drop in performance once caches are too small to hold their working sets.
+Thus, these applications are likely to be bandwidth-bound at main memory unless some significant changes are made to the memory hierarchy.
+
+Unfortunately, there are few simple techniques to improve bandwidth efficiency of a system.
+While there is some work in this area~\cite{BW-Wall}, this is insufficient for the large bandwidth requirements of some throughput computing applications.
+So today¡¯s systems that tout good performance for throughput computing applications (e.g., nVidia¡¯s Tesla) do so by providing large main memory bandwidths via the use of GDDR rather than improving bandwidth efficiency.
+However, GDDR has fairly strict capacity limits and is much more power hungry than conventional DRAM modules.
+This reduces its desirability for throughput computing, and makes it a bad choice for general purpose systems trying to improve their throughput computing performance.
+
+Emerging memory technologies such as Magnetoresistive Random Access Memory~(MRAM), Phase-Change Memory~(PCM), Resistive Random Access Memory~(RRAM),
+embedded DRAM, have shown potential to to be used as on-chip caches and the main memory~\cite{Sun:2009:MRAM-L2-CMP}.
+Therefore, we explore how to enhance the memory hierarchy from the bandwidth point of view.
+
+% -----------------------------------------------------------------------------
+
+\section{Emerging Memory Technologies}
+
+We first review the technology background three types of non-volatile memories, which are STT-RAM, PCRAM, and RRAM.
+
+\subsection{STT-RAM}
+
+\subsection{PCRAM}
+
+\subsection{RRAM}
+
+% -----------------------------------------------------------------------------
+\section{Memory Technology Pre-Exploration}
+
+We use our modified CACTI~\cite{CACTI} to estimate the bandwidths provided
+by different memory technologies.
 
@@
 
+\section{Experiments}
+
+ We will also use
+Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
+design method. The benchmarks will be selected from bandwidth-bound
+applications, which possibly include benchmarks from SPEC
+CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}
+
+\subsection{Experimental Setup}
+
+\subsection{Results}
+
+\section{Conclusion and Future Work}
+
 In addition, we would like to put an upper bound to the total dynamic energy

### 2010-11-17 trunk/paper.tex
some figure and paragraph updated

@@
 \subsection{STT-RAM}
+STT-RAM uses the magnetic property of the material and uses Magnetic Tunnel Junction (MTJ) as its binary storage.
+As shown in Fig.~\ref{fig:mram_cell}, MTJ contains two ferromagnetic layers and one tunnel barrier layer.
+The direction of one ferromagnetic layer is fixed, which is called the reference layer, while the direction of the other one can be changed by passing a driving current, which is called the free layer.
+The relative magnetization direction of two ferromagnetic layers determines the resistance of MTJ.
+If two ferromagnetic layers have the same directions, the resistance of MTJ is low, indicating a ``0" state; if two layers have different directions, the resistance of MTJ is high, indicating a ``1" state.
+
+\begin{figure*}[htbp]
+% The "!" means to maintain the aspect ratio.
+\centering
+\includegraphics[width=3in]{figures/mram_cell}
+\caption{Demonstration of a MRAM cell. (a) Structural view. (b) Schematic view.}
+\label{fig:mram_cell}
+\end{figure*}
 
@@
 
+PCRAM uses chalcogenide-based material to storage informations.
+The chalcogenide-based materials in recent PCRAM research are usually alloys of germanium, antimony, and tellurium (GeSbTe, or GST), which can be switched between a crystalline phase (SET or ``1" state) and an amorphous phase (RESET or ``0" state) with the application of heat.
+The crystalline phase shows high optical reflectivity and low electrical resistivity, while the amorphous phase is characterized by low reflectivity and high resistivity.
+
+\begin{figure*}[htbp]
+% The "!" means to maintain the aspect ratio.
+\centering
+\includegraphics[width=3in]{figures/GST}
+\caption{The schematic view of a PCRAM cell with MOSFET selector transistor (BL=Bitline, WL=Wordline, SL=Sourceline}
+\label{fig:GST}
+\end{figure*}
+
 \subsection{RRAM}
+
+\begin{figure*}[htbp]
+% The "!" means to maintain the aspect ratio.
+\centering
+\includegraphics[width=3in]{figures/memristor}
+\caption{The conceptual view of the structure of memristor cells.}
+\label{fig:memristor}
+\end{figure*}
+
+Memristor, a portmanteau of ``memory resistor'', is a generalized resistance that maintains a functional relationship between the time integrals of current and voltage.
+Memristor was first theoretically predicted by Chua in 1971~\cite{memristor:chua} as the fourth fundamental circuit element from the completeness of relations between the four basic circuit variables, namely, current, voltage, charge, and flux-linkage.
+The first memristor practical demonstration was presented by Williams \emph{et al.} in 2008~\cite{memristor:missing}.
+Fig.~\ref{fig:memristor} shows a conceptual view of the memristor structure~\cite{memristor:missing}.
+The top electrode and bottom electrode are two metal nanowires on platinum, and the thin titanium dioxide film is sandwiched by the electrodes.
 
@@
 \bibliographystyle{latex8}
-\bibliography{./bib/ref}
+\bibliography{./bib/ref,./bib/reram}
 

### 2010-11-17 trunk/paper.tex
description of NVsim added

@@
 \section{Memory Technology Pre-Exploration}
-
-We use our modified CACTI~\cite{CACTI} to estimate the bandwidths provided
-by different memory technologies.
-
-In this section, we describe the experimental results we have obtained so far
-using a modified version of CACTI~\cite{CACTI}. \vspace{-0.5in}
+Many modeling tools have been developed during the last decade to enable system-level design exploration for SRAM- or DRAM-based cache and memory design.
+For example, CACTI~\cite{CACTI} is a tool that has been widely used in the computer architecture community to estimate the speed, power, and area of SRAM and DRAM caches.
+In addition, CACTI has also been extended to evaluate the performance, power, and area for STT-RAM~\cite{CACTI:DAC08:Dong}, PCRAM~\cite{CACTI:GLSVLSI08:Mangalagiri,CACTI:PCRAMsim}, and NAND flash~\cite{CACTI:DATE10:Mohan}.
+However, as CACTI is originally designed to model SRAM-based cache, some of its fundamental assumptions do not match the actual NVM circuit implementation, and thereby these CACTI-like estimation tools do not model the NVM array organization in the exact way that the chip is fabricated.
+In this section, we use \emph{NVSim}, a circuit-level model for NVM performance, energy, and area estimation, which supports various NVM technologies including STT-RAM, PCRAM, RRAM, and conventional NAND flash.
 
@@
 \bibliographystyle{latex8}
-\bibliography{./bib/ref,./bib/reram}
+\bibliography{./bib/ref,./bib/reram,./bib/cacti}
 

### 2010-11-18 paper.tex
Grey figure updated

@@
 % -----------------------------------------------------------------------------
+\section{Motivation}
 
-\section{Experimental Setup}
+Throughput computing involves performing a huge number of calculations with a large amount of parallelism.
+Throughput computing applications span many domains and are already critical on a variety of platforms from high-performance computing (HPC) machines to commercial servers to client machines.
 
-We will use our modified CACTI~\cite{CACTI} to estimate the bandwidths provided
-by different memory technologies. We will also use
-Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
-design method. The benchmarks will be selected from bandwidth-bound
-applications, which possibly include benchmarks from SPEC
-CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}
+Systems designed to perform well on throughput computing applications achieve high performance by exploiting their inherent parallelism.
+These systems support large numbers of threads and/or use wide SIMD execution.
+This puts a lot of pressure on the memory system.
+Fortunately, memory latency is typically not a bottleneck since the latency can be hidden via multithreading or hardware prefetching, since the data access patterns are relatively simple.
+On the other hand, bandwidth is a potential bottleneck.
 
-\section{Experiments}
+Many throughput computing applications have inherently large working sets (i.e., tens to hundreds of MB).
+These are unlikely to fit in conventional on-die SRAM caches for the foreseeable future.
+Further, unlike more traditional workloads (e.g., those similar to SPEC or TPC benchmarks), many throughput computing applications see a sharp drop in performance once caches are too small to hold their working sets.
+Thus, these applications are likely to be bandwidth-bound at main memory unless some significant changes are made to the memory hierarchy.
 
-In this section, we describe the experimental results we have obtained so far
-using a modified version of CACTI~\cite{CACTI}. \vspace{-0.5in}
+Unfortunately, there are few simple techniques to improve bandwidth efficiency of a system.
+While there is some work in this area~\cite{BW-Wall}, this is insufficient for the large bandwidth requirements of some throughput computing applications.
+So today¡¯s systems that tout good performance for throughput computing applications (e.g., nVidia¡¯s Tesla) do so by providing large main memory bandwidths via the use of GDDR rather than improving bandwidth efficiency.
+However, GDDR has fairly strict capacity limits and is much more power hungry than conventional DRAM modules.
+This reduces its desirability for throughput computing, and makes it a bad choice for general purpose systems trying to improve their throughput computing performance.
+
+Emerging memory technologies such as Magnetoresistive Random Access Memory~(MRAM), Phase-Change Memory~(PCM), Resistive Random Access Memory~(RRAM),
+embedded DRAM, have shown potential to to be used as on-chip caches and the main memory~\cite{Sun:2009:MRAM-L2-CMP}.
+Therefore, we explore how to enhance the memory hierarchy from the bandwidth point of view.
+
+% -----------------------------------------------------------------------------
+
+\section{Emerging Memory Technologies}
+
+We first review the technology background three types of non-volatile memories, which are STT-RAM, PCRAM, and RRAM.
+
+\subsection{STT-RAM}
+STT-RAM uses the magnetic property of the material and uses Magnetic Tunnel Junction (MTJ) as its binary storage.
+As shown in Fig.~\ref{fig:mram_cell}, MTJ contains two ferromagnetic layers and one tunnel barrier layer.
+The direction of one ferromagnetic layer is fixed, which is called the reference layer, while the direction of the other one can be changed by passing a driving current, which is called the free layer.
+The relative magnetization direction of two ferromagnetic layers determines the resistance of MTJ.
+If two ferromagnetic layers have the same directions, the resistance of MTJ is low, indicating a ``0" state; if two layers have different directions, the resistance of MTJ is high, indicating a ``1" state.
 
@@
 \centering
-\includegraphics[width=3in]{figures/read-bw}
-\includegraphics[width=3in]{figures/write-bw}\\
+\includegraphics[width=3in]{figures/mram_cell}
+\caption{Demonstration of a MRAM cell. (a) Structural view. (b) Schematic view.}
+\label{fig:mram_cell}
+\end{figure*}
 
-\vspace*{-1in}
-\hspace{0.02in}
-\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
-\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}
+\subsection{PCRAM}
 
+PCRAM uses chalcogenide-based material to storage informations.
+The chalcogenide-based materials in recent PCRAM research are usually alloys of germanium, antimony, and tellurium (GeSbTe, or GST), which can be switched between a crystalline phase (SET or ``1" state) and an amorphous phase (RESET or ``0" state) with the application of heat.
+The crystalline phase shows high optical reflectivity and low electrical resistivity, while the amorphous phase is characterized by low reflectivity and high resistivity.
+
+\begin{figure*}[htbp]
+% The "!" means to maintain the aspect ratio.
+\centering
+\includegraphics[width=3in]{figures/GST}
+\caption{The schematic view of a PCRAM cell with MOSFET selector transistor (BL=Bitline, WL=Wordline, SL=Sourceline}
+\label{fig:GST}
+\end{figure*}
+
+\subsection{RRAM}
+
+\begin{figure*}[htbp]
+% The "!" means to maintain the aspect ratio.
+\centering
+\includegraphics[width=3in]{figures/memristor}
+\caption{The conceptual view of the structure of memristor cells.}
+\label{fig:memristor}
+\end{figure*}
+
+Memristor, a portmanteau of ``memory resistor'', is a generalized resistance that maintains a functional relationship between the time integrals of current and voltage.
+Memristor was first theoretically predicted by Chua in 1971~\cite{memristor:chua} as the fourth fundamental circuit element from the completeness of relations between the four basic circuit variables, namely, current, voltage, charge, and flux-linkage.
+The first memristor practical demonstration was presented by Williams \emph{et al.} in 2008~\cite{memristor:missing}.
+Fig.~\ref{fig:memristor} shows a conceptual view of the memristor structure~\cite{memristor:missing}.
+The top electrode and bottom electrode are two metal nanowires on platinum, and the thin titanium dioxide film is sandwiched by the electrodes.
+
+% -----------------------------------------------------------------------------
+\section{Memory Technology Pre-Exploration}
+Many modeling tools have been developed during the last decade to enable system-level design exploration for SRAM- or DRAM-based cache and memory design.
+For example, CACTI~\cite{CACTI} is a tool that has been widely used in the computer architecture community to estimate the speed, power, and area of SRAM and DRAM caches.
+In addition, CACTI has also been extended to evaluate the performance, power, and area for STT-RAM~\cite{CACTI:DAC08:Dong}, PCRAM~\cite{CACTI:GLSVLSI08:Mangalagiri,CACTI:PCRAMsim}, and NAND flash~\cite{CACTI:DATE10:Mohan}.
+However, as CACTI is originally designed to model SRAM-based cache, some of its fundamental assumptions do not match the actual NVM circuit implementation, and thereby these CACTI-like estimation tools do not model the NVM array organization in the exact way that the chip is fabricated.
+In this section, we use \emph{NVSim}, a circuit-level model for NVM performance, energy, and area estimation, which supports various NVM technologies including STT-RAM, PCRAM, RRAM, and conventional NAND flash.
+
+\begin{figure*}[htbp]
+% The "!" means to maintain the aspect ratio.
+\centering
+\includegraphics[width=6in]{figures/RAM_bandwidth}
 \caption{Read and write bandwidths provided by different memory technologies.
@@
 \centering
-\includegraphics[width=2in]{figures/read-energy}
-\includegraphics[width=2in]{figures/write-energy}
-\includegraphics[width=2in]{figures/total-energy}\\
+\includegraphics[width=6in]{figures/RAM_energy}
+\caption{Dynamic energy consumption with the provided bandwidths of different
+  memory technologies. (a) Dynamic Energy with 20\% write.  (b) Dynamic Energy with 50\% write. }
+\label{fig:memory-energy}
+\end{figure*}
 
-\vspace*{-0.5in}
-\hspace{0.02in}
-\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
-\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}
-\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (c)}}
-
-\caption{Dynamic energy consumption with the provided bandwidths of different
-  memory technologies. (a) Read energy of different memory technologes. (b)
-  Write energy of different memory technologies. (c) Total energy consumption
-  (with 10\% write) of different memory technologies.}
-\label{fig:memory-energy}
+\begin{figure*}[htbp]
+% The "!" means to maintain the aspect ratio.
+\centering
+\includegraphics[width=6in]{figures/RAM_latency}
+\caption{Latency of different memory technologies. (a) Read Latency. (b) Write Latency.}
+\label{fig:latency}
 \end{figure*}
@@
 \bibliographystyle{latex8}
-\bibliography{./bib/ref}
+\bibliography{./bib/ref,./bib/reram,./bib/cacti}
 

### 2010-11-18 paper.tex
experiment part added

@@
 \usepackage{wrapfig}
+\usepackage{multirow}
 
@@
 
- We will also use
-Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
-design method. The benchmarks will be selected from bandwidth-bound
-applications, which possibly include benchmarks from SPEC
-CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}
+Based on the parameters of different cache configurations collected from our
+modified version of CACTI~\cite{CACTI}, we evaluate both pure SRAM-based and
+hybrid cache hierarchy designs. In this section, we show how bandwidth-aware
+memory hierarchy design method help with improving the system performance.
 
@@
 
+We use Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator in our
+experiments. It is configured to model an eight-core CMP. Each core is in-order, and
+is similar to UltraSPARC III architecture. The frequency of each core is set to
+be 1GHz. Table~\ref{table:cmp-config} lists the detailed parameters of the
+baseline. Our baseline contains 8 private L1 instruction and data caches
+respectively. Since we only evaluate shared cache hierarchies, each of L1 cache
+is fixed to be SRAM-based, and of 64KB capacity. As regard to lower level
+caches, we evaluate two cases, pure SRAM-based and hybrid caches with various
+memory technologies, including SRAM, MRAM, RRAM, and eDRAM. By evaluating both
+cases, we would like to find out optimal cache design that leads to the peak
+performance, i.e., the number of cache levels that is reqired, memory technology
+used for each level, and capacity of each level.
+
+\begin{table}[t]
+  \centering  
+  \caption{Baseline CMP configuration.}
+  \vspace{0.1in}
+  \begin{tabular}{l|c}
+    \hline
+    \multicolumn{2}{|l|}{\textbf{Core}}\\
+    \hline
+    No. of cores & 8\\
+    \hline
+    Frequency & 1GHz\\
+    \hline
+    Core architecture & in-order, 14-stage pipline, 2 ALUs, 2 FPUs\\
+    \hline
+    \multicolumn{2}{|l|}{\textbf{Memory}}\\
+    \hline    
+    \multirow{2}{*}{Private caches} & L1 I-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
+                                    & L1 D-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
+    \hline
+    \multirow{2}{*}{Shared caches} & Case 1: Pure SRAM, 64B line, 8-way, write-back \\
+                                   & Case 2: Hybrid~(SRAM, MRAM, RRAM, eDRAM), 64B line, 8-way, write-back\\
+    \hline    
+    Main memory & 4GB\\
+    \hline    
+  \end{tabular}
+  \label{table:cmp-config}
+\end{table}
+
+The benchmarks are selected from PARSEC benchmark suite~\cite{PARSEC:2008} with
+multithreaded programs, which focus on emerging workloads that are designed to
+be representative of next-generation shared-memory programs for CMPs. Since the
+performance of different memory technologies are closely related to read and
+write intensities, we selected some workloads that vary in the average numbers
+of L2 cache read per thousand instructions~(RPKI) and write per thousand
+instructions~(WPKI), which are listed in Table~\ref{table:benchmarks}.
+
+\begin{table}[htbp!]
+  \centering  
+  \caption{Characteristics of selected benchmarks.}
+  \vspace{0.1in}
+  \begin{tabular}{l|c|c||l|c|c}
+    \hline    
+    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI} & 
+    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI}\\
+    \hline
+    blackscholes & 22.8 & 61.7 & ferret & 5.7 & 173.4\\
+    \hline 
+    bodytrack & 5.4 & 139.5 & fluidanimate & 2.6 & 70.1\\
+    \hline
+    canneal & 5.4 & 29.3 & streamcluster & 17.3 & 16.9\\
+    \hline
+    facesim & 6.0 & 102.4 & swaptions & 2.6 & 121.4\\
+    \hline
+  \end{tabular}
+  \label{table:benchmarks}
+\end{table}
+
 \subsection{Results}
 
+We evaluate various possible configurations with shared caches by simulation,
+i.e., with possible numbers of levels, memory technologies, and cache capacities
+for each cache level. We consider SRAM and eDRAM as the possible memory
+technologies to implement L2 and L3 caches. Both MRAM and RRAM have higher write
+latency than SRAM. Furthermore, the endurance of RRAM is too low to be used as
+lower level caches. Consequently, these two memory technologies are only
+considered to be used as the last level cache.
+
+In the first set of experiments, we evaluate the system performance with
+two-level shared caches. Figure~\ref{fig:thput-l2l3} shows the system throughput
+with all the benchmarks. It is illustrated that implementing hybrid cache with
+eDRAM as the L3 cache helps with performance among most of the benchmarks. Such
+performance improvement is more than 10\% in Figure~\ref{fig:thput-l2l3}(a). It
+is indicated by our memory technology exploration that eDRAM shows more latency
+benefits with larger capacities. As a result, a larger eDRAM-based L3 cache
+leads to more performance improvement to the pure SRAM implementation, as
+illustrated by Figure~\ref{fig:thput-l2l3}(b). However, hybrid cache does not
+always improve the performance, as shown in Figure~\ref{fig:benchmark-l2l3}.
+With the benchmark \emph{canneal}, hybrid cache configurations outperform the
+pure SRAM implementation among overall range of various capacities. With the
+benchmark \emph{ferret}, however, pure SRAM implementation results in higher
+performance than hybrid cache implementation with the same capacity
+configuration.
+
+\begin{figure*}[t] 
+\centering
+\includegraphics[width=3in]{figures/thput-l2l3-1}\hspace{0.2in}
+\includegraphics[width=3in]{figures/thput-l2l3-2}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+\caption{Performance comparison with two-level caches among various benchmarks.
+  (a) The L3 cache capacity is 16MB. (b) The L3 cache capacity is 32MB.}
+\label{fig:thput-l2l3}
+\end{figure*}
+
+\begin{figure*}[t] 
+\centering
+\includegraphics[width=3in]{figures/canneal-l2l3}\hspace{0.2in}
+\includegraphics[width=3in]{figures/ferret-l2l3}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+\caption{Performance comparison with two-level caches among various cache
+  capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
+\label{fig:benchmark-l2l3}
+\end{figure*}
+
+In the second set of experiments, we evaluate the system performance with
+three-level shared caches. When evaluating hybrid cache configurations, we
+consider implementing the last level cache by MRAM and RRAM memory technologies,
+since the data transaction intensity is relatively low in the last level cache.
+Figure~\ref{fig:thput-l2l3l4} shows the results.
+Figure~\ref{fig:thput-l2l3l4}(a) illustrated that the performance more than 15\%
+higher on average with a last level cache implemented by MRAM than pure SRAM
+implementations. More performance improvement is obtained by increasing the last
+level cache capacity, as shown in Figure~\ref{fig:thput-l2l3l4}(b).
+Figure~\ref{fig:benchmark-l2l3l4} compares performance with pure SRAM and hybrid
+cache implementations with different capacities. In this case, hybrid cache
+implementation leads to higher performance in both benchmarks. The results of
+the two sets of experiments are illustrated together in
+Figure~\ref{fig:benchmark-overall}. An interesting observation is that hybrid
+cache implementation shows high performance improvement with one of the
+applications \emph{canneal}, whereas as little improvement with the other
+\emph{ferret}. The reason is that write latency of both MRAM and RRAM is higher
+than SRAM when capacity is less than 1GB. Performance of applications with large
+number of writes, such as \emph{ferret} is therefore not benefit from hybrid
+cache design.
+
+\begin{figure*}[t] 
+\centering
+\includegraphics[width=3in]{figures/thput-l2l3l4-1}\hspace{0.2in}
+\includegraphics[width=3in]{figures/thput-l2l3l4-2}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+
+\caption{Performance comparison with three-level caches among various
+  benchmarks. (a) The L3 cache capacity is 16MB, and the L4 cache capacity is
+  512MB. (b) The L3 cache capacity is 32MB, and the L4 cache capacity is 1GB.}
+\label{fig:thput-l2l3l4}
+\end{figure*}
+
+\begin{figure*}[t] 
+\centering
+\includegraphics[width=3in]{figures/canneal-l2l3l4}\hspace{0.2in}
+\includegraphics[width=3in]{figures/ferret-l2l3l4}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+\caption{Performance comparison with three-level caches among various cache
+  capacity. (a) The canneal benchmark. (b) The ferret benchmark..}
+\label{fig:benchmark-l2l3l4}
+\end{figure*}
+
+\begin{figure*}[t] 
+\centering
+\includegraphics[width=3in]{figures/canneal}\hspace{0.2in}
+\includegraphics[width=3in]{figures/ferret}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+
+\caption{Performance comparison with both two-level and three-level caches among
+  various cache capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
+\label{fig:benchmark-overall}
+\end{figure*}
+
+Finally, we evaluate all the possible cache hierarchy configurations with
+exhaustive simulations. The optimal cache configurations of each benchmark is
+shown in Table~\ref{table:results}. Since the characteristics vary among the
+benchmarks, the optimal cache configurations are different among each of
+benchmarks. Applications with high write intensities, such as \emph{bodytrack},
+\emph{ferret}, and \emph{swaptions} tend to favor SRAM-based caches. Other
+applications benefit more from hybrid cache implementation in terms of
+performance.
+
+\begin{table}[htbp!]
+  \centering  
+  \caption{Optimal cache configurations for each benchmark.}
+  \vspace{0.1in}
+  \begin{tabular}{|l|c|c|c|}
+    \hline    
+    \textbf{Benchmarks} & \textbf{L2 Configuration} & \textbf{L3 Configuration} & \textbf{L4 Configuration}\\
+    \hline
+    blackscholes & SRAM, 8MB & eDRAM, 32MB & RRAM, 1GB\\
+    \hline 
+    bodytrack & SRAM, 8MB & eDRAM, 128MB & -\\
+    \hline
+    canneal & SRAM, 8MB & eDRAM, 512MB & MRAM, 1GB\\
+    \hline
+    facesim & SRAM, 8MB & eDRAM, 32MB & MRAM, 1GB\\
+    \hline
+    ferret & SRAM, 8MB & eDRAM, 32MB & -\\
+    \hline
+    fluidanimate & SRAM, 8MB & eDRAM, 1GB & -\\
+    \hline
+    streamcluster & SRAM, 16MB & eDRAM, 128MB & MRAM, 512MB\\
+    \hline
+    swaptions & SRAM, 8MB & SRAM, 32MB & SRAM, 1GB\\
+    \hline
+  \end{tabular}
+  \label{table:results}
+\end{table}
+
 \section{Conclusion and Future Work}
+In this project, we explore performance and energy characteristics of various
+emerging memory technolgies. Based on our exploration, we evaluate the shared
+cache hierarchy design of CMPs optimized for performance by filling the off-chip
+memory bandwidth gap. According to our evaluation, SRAM-based caches leads to
+reasonable performance with write-intensive applications. Hybrid cache design
+help with performance with other types of benchmarks.
 
-In addition, we would like to put an upper bound to the total dynamic energy
-consumption of the memory hierarchy. Figure~\ref{fig:memory-energy} shows the
-estimation of dynamic power consumptions of different memory technologies with
-the provided bandwidths. In figure~\ref{fig:memory-energy}(c), we estimate the total
-dynamic energy with 10\% write activities.
+Furthermore, a variety of continuing studies is remained to be explored related
+to bandwidth-aware memory hierarchy design. First of all, we only examine cache
+hierarchy design currently. It is necessary to extend our exploration to the
+overall memory hierarchy design. On the other hand, we do not consider energy
+efficiency in this project. We would like to put some power constraints to the
+memory hierarchy, and examine the energy efficiency of various memory hierarchy
+configurations.
 

### 2010-11-18 paper.tex
combined version

@@
 \usepackage{wrapfig}
+\usepackage{enumerate}
 \usepackage{multirow}
@@
 
+First of all, we estimate the read and write bandwidths that can be provided by different memory technologies.
+Figure~\ref{fig:memory-bw} shows the results, with both x- and y-values in \emph{log} scale.
+The figure illustrates both the
+provided read and write bandwidth as a function of memory capacity.
+Each of the memory technologies actually provide nearly the same read bandwidths, as is shown in figure~\ref{fig:memory-bw}(a).
+On the other hand, a straight forward observation from figure~\ref{fig:memory-bw}(b) is that the write bandwidth varies among different memory technologies.
+The shape of the SRAM write bandwidth curve is very similar to the read bandwidth curve.
+The write bandwidth curves of the other three memory technologies appear to be very different.
+The reason is that write latencies of the three non-volatile memories are much higher than read latencies.
+Another observation from figure~\ref{fig:memory-bw}(b) is that the curves cross to each other at different locations.
+
 \begin{figure*}[htbp]
@@
 \label{fig:memory-energy}
+\includegraphics[width=6in]{figures/RAM_latency}
+\caption{Latency of different memory technologies. (a) Read Latency. (b) Write Latency.}
+\label{fig:memory-latency}
 \end{figure*}
 
-\begin{figure*}[htbp]
-% The "!" means to maintain the aspect ratio.
-\centering
-\includegraphics[width=6in]{figures/RAM_latency}
-\caption{Latency of different memory technologies. (a) Read Latency. (b) Write Latency.}
-\label{fig:latency}
-\end{figure*}
+Being aware that the read and write latencies of NVM are asymmetric, we consider the read latency at first. In NVsim, we divide the entire cache read latency into the following components:
+\begin{itemize}
+\item (a)	H-tree input delay
+\item (b)	Decoder + word-line delay
+\item (c)	Bit-line delay
+\item (d)	Sense Amplifier delay
+\item (e)	Comparator Delay (for tag part only)
+\item (f)	H-tree output delay
+\end{itemize}
+H-tree latency (a) and (f) are mainly determined by the RC delay of global wires, which is positive proportional to the area of memory macro.
+Sensing delay (d) is affected by the read noise margin of memory cell that is affected by off/on resistance ratio.
 
-First of all, we estimate the read and write bandwidths that can be provided by
-different memory technologies. Figure~\ref{fig:memory-bw} shows the results,
-with both x- and y-values in \emph{log} scale. The figure illustrates both the
-provided read and write bandwidth as a function of memory capacity. Each of the
-memory technologies actually provide nearly the same read bandwidths, as is
-shown in figure~\ref{fig:memory-bw}(a). On the other hand, a straight forward
-observation from figure~\ref{fig:memory-bw}(b) is that the write bandwidth
-varies among different memory technologies. The shape of the SRAM write
-bandwidth curve is very similar to the read bandwidth curve. The write bandwidth
-curves of the other three memory technologies appear to be very different. The
-reason is that write latencies of the three non-volatile memories are much
-higher than read latencies. Another observation from
-figure~\ref{fig:memory-bw}(b) is that the curves cross to each other at
-different locations. Based on these two observations, we would like to design a
+Based on these two observations, we would like to design a
 bandwidth-aware hybrid memory hierarchy, which always provides the high memory
@@
 
-Based on the parameters of different cache configurations collected from our
-modified version of CACTI~\cite{CACTI}, we evaluate both pure SRAM-based and
-hybrid cache hierarchy designs. In this section, we show how bandwidth-aware
-memory hierarchy design method help with improving the system performance.
+ We will also use
+Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
+design method. The benchmarks will be selected from bandwidth-bound
+applications, which possibly include benchmarks from SPEC
+CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}
 

### 2010-11-18 paper.tex
abstract

@@
 \begin{abstract}
-\boldmath
+  \boldmath Chip-multiprocessor~(CMP) is a promising solution for high
+  performance computing systems. However, the limited increases in main memory
+  bandwidth results in a growing bandwidth gap between the processor cores and
+  the main memory, which becomes a potential obstacles to system performance
+  improvement. In this project, we explore bandwidth-aware memory hierarchy
+  design with various emerging memory technologies. Based on the observations
+  obtained from our pre-exploration of various emerging memory technologies, we
+  evaluate systerm performance obtained with various cache hierarchy
+  configurations. With different applications, we evaluate the optimal shared
+  cache configurations in a CMP system, in terms of the number of shared cache
+  levels, memory technology and capacity of each level. We show that cache
+  hierarchy with hybrid memory technologies benefits system performance with
+  non-write-intensive applications.
 
@@
 multiple data~(SIMD) execution, which puts a lot of pressure on the memory
-system. Memory latency is typically not a bottleneck, since the latency can be
-hidden via multi-threading or hardware prefetching. To this end, bandwidth becomes a
-potential bottleneck. A high rate of computing often brings in a high rate of
-data transitions. In some cases, the working set of an application fits in the
-on-die caches, which can typically provide sufficient bandwidth to keep up with
-the processing cores. However, if the working set does not fit in the on-die
-caches, the main memory needs to provide much of the data. Since the bandwidth
-of off-chip main memory is quite limited, applications with such working sets
-are potentially bandwidth-bound. Therefore, it is crucial to design the memory
-hierarchy to overcome the bandwidth limitation.
+system. Typically, memory latency is not a bottleneck, since the latency can be
+hidden via multi-threading or hardware prefetching. To this end, bandwidth
+becomes a potential bottleneck. A high rate of computing often brings in a high
+rate of data transitions. In some cases, the working set of an application fits
+in the on-die caches, which can typically provide sufficient bandwidth to keep
+up with the processing cores. However, if the working set does not fit in the
+on-die caches, the main memory needs to provide much of the data. Since the
+bandwidth of off-chip main memory is quite limited, applications with such
+working sets are potentially bandwidth-bound. Therefore, it is crucial to design
+the memory hierarchy to overcome the bandwidth limitation.
 
@@
 
- We will also use
-Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
-design method. The benchmarks will be selected from bandwidth-bound
-applications, which possibly include benchmarks from SPEC
-CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}
+Based on the parameters of different cache configurations collected from our
+modified version of CACTI~\cite{CACTI}, we evaluate both pure SRAM-based and
+hybrid cache hierarchy designs. In this section, we show how bandwidth-aware
+memory hierarchy design method help with improving the system performance.\vspace{0.15in}
 
@@
     \hline
-    \multicolumn{2}{|l|}{\textbf{Core}}\\
+    \multicolumn{2}{l}{\textbf{Core}}\\
     \hline
@@
     \hline
-    \multicolumn{2}{|l|}{\textbf{Memory}}\\
+    \multicolumn{2}{l}{\textbf{Memory}}\\
     \hline
@@
 of L2 cache read per thousand instructions~(RPKI) and write per thousand
-instructions~(WPKI), which are listed in Table~\ref{table:benchmarks}.
+instructions~(WPKI), which are listed in
+Table~\ref{table:benchmarks}.\vspace{0.3in}
 
-\begin{table}[htbp!]
+\begin{table}[t]
   \centering
@@
   \label{table:benchmarks}
+\vspace{0.5in}
 \end{table}

### 2010-11-18 trunk/paper.tex
version 1.0

@@
 \usepackage{wrapfig}
+\usepackage{enumerate}
+\usepackage{multirow}
 
@@
 
-Many of modern chip-multiprocessors are designed to perform well on various
-applications to achieve high performance by exploiting their inherent
-parallelism. Such systems support large number of threads and single instruction
-multiple data~(SIMD) execution, which puts a lot of pressure on the memory
-system. Memory latency is typically not a bottleneck, since the latency can be
-hidden via multi-threading or hardware prefetching. To this end, bandwidth becomes a
-potential bottleneck. A high rate of computing often brings in a high rate of
-data transitions. In some cases, the working set of an application fits in the
-on-die caches, which can typically provide sufficient bandwidth to keep up with
-the processing cores. However, if the working set does not fit in the on-die
-caches, the main memory needs to provide much of the data. Since the bandwidth
-of off-chip main memory is quite limited, applications with such working sets
-are potentially bandwidth-bound. Therefore, it is crucial to design the memory
-hierarchy to overcome the bandwidth limitation.
+Many of modern chip-multiprocessors are designed to perform well on various applications to achieve high performance by exploiting their inherent parallelism.
+Such systems support large number of threads and single instruction multiple data~(SIMD) execution, which puts a lot of pressure on the memory system.
+Typically, memory latency is not a bottleneck, since the latency can be hidden via multi-threading or hardware prefetching.
+To this end, bandwidth becomes a potential bottleneck.
+A high rate of computing often brings in a high rate of data transitions.
+In some cases, the working set of an application fits in the on-die caches, which can typically provide sufficient bandwidth to keep up with the processing cores.
+However, if the working set does not fit in the on-die caches, the main memory needs to provide much of the data.
+Since the bandwidth of off-chip main memory is quite limited, applications with such working sets are potentially bandwidth-bound.
+Therefore, it is crucial to design the memory hierarchy to overcome the bandwidth limitation.
 
-It is known that performance improvement of a computing system can be achieved
-via multiple memory levels. Adding an extra level of memory, however, can also
-help alleviate the bandwidth bottleneck of off-chip memory. Specifically, we
-would like to examine (1) the number of levels in the optimal memory hierarchy,
-(2) the appropriate memory technology of each level, and (3) the capacity and
-bandwidth of each level. We will also explore the energy-efficiency of the
-memory hierarchy design constrain, and the memory hierarchy design within a
-fixed power budge.\vspace{0.15in}
+It is known that performance improvement of a computing system can be achieved via multiple memory levels.
+Adding an extra level of memory, however, can also help alleviate the bandwidth bottleneck of off-chip memory.
+Specifically, we examine (1) the number of levels in the optimal memory hierarchy, (2) the appropriate memory technology of each level, and (3) the capacity and bandwidth of each level.
+We will also explore the energy-efficiency of the memory hierarchy design constrain, and the memory hierarchy design within a fixed power budge.
+
+Emerging memory that provides fast random access, high storage density, and non-volatility within one memory technology becomes possible  due to the emergence of  various new non-volatile memory~(NVM) technologies, such as spin-torque-transfer memory~(STT-RAM), phase-change memory~(PCRAM), and resistive memory~(RRAM).
+Our goal in this project to to leverage these NVM in devising a memory system with multiple layers of the existing memory hierarchy for modern computers, to supply a wide spectrum of final products that cover from a highly latency-optimized microprocessor cache to a highly density-optimized secondary storage device.
+
+The report is organized as follows.
+First, we present the motivation of design a memory system with better memory bandwidth.
+Second, we make a quick summary of current emerging memory technologies.
+Then we use our NVsim, a circuit-level NVM model to evaluate the read latency, write latency and dynamic energy of different memory technologies.
+After that, we perform the cycle-accurate architectural simulation to find the best memory system design for different applications.
+Finally we conclude our work and propose potential future work.
 
@@
 \centering
-\includegraphics[width=3in]{figures/read-bw}
-\includegraphics[width=3in]{figures/write-bw}\\
-
-\vspace*{-1in}
-\hspace{0.02in}
-\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
-\makebox[3in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}
-
+\includegraphics[width=6in]{figures/RAM_bandwidth}
 \caption{Read and write bandwidths provided by different memory technologies.
@@
 \centering
-\includegraphics[width=2in]{figures/read-energy}
-\includegraphics[width=2in]{figures/write-energy}
-\includegraphics[width=2in]{figures/total-energy}\\
-
-\vspace*{-0.5in}
-\hspace{0.02in}
-\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (a)}}
-\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (b)}}
-\makebox[2in][l]{\textcolor [rgb]{0,0,0} {\bf (c)}}
-
+\includegraphics[width=6in]{figures/RAM_latency}
+\caption{Latency of different memory technologies. (a) Read Latency. (b) Write Latency.}
+\label{fig:memory-latency}
+\includegraphics[width=6in]{figures/RAM_energy}
 \caption{Dynamic energy consumption with the provided bandwidths of different
-  memory technologies. (a) Read energy of different memory technologes. (b)
-  Write energy of different memory technologies. (c) Total energy consumption
-  (with 10\% write) of different memory technologies.}
+  memory technologies. (a) Dynamic Energy with 20\% write.  (b) Dynamic Energy with 50\% write. }
 \label{fig:memory-energy}
@@
 
-First of all, we estimate the read and write bandwidths that can be provided by
-different memory technologies. Figure~\ref{fig:memory-bw} shows the results,
-with both x- and y-values in \emph{log} scale. The figure illustrates both the
-provided read and write bandwidth as a function of memory capacity. Each of the
-memory technologies actually provide nearly the same read bandwidths, as is
-shown in figure~\ref{fig:memory-bw}(a). On the other hand, a straight forward
-observation from figure~\ref{fig:memory-bw}(b) is that the write bandwidth
-varies among different memory technologies. The shape of the SRAM write
-bandwidth curve is very similar to the read bandwidth curve. The write bandwidth
-curves of the other three memory technologies appear to be very different. The
-reason is that write latencies of the three non-volatile memories are much
-higher than read latencies. Another observation from
-figure~\ref{fig:memory-bw}(b) is that the curves cross to each other at
-different locations. Based on these two observations, we would like to design a
-bandwidth-aware hybrid memory hierarchy, which always provides the high memory
-bandwidth with the given capacity. For example, we probably would like to use
-1) SRAM when the memory capacity is under 256MB, 2) MRAM or RRAM between
-256MB and 1GB (if only bandwidth is considered), and 3) PCRAM when the memory
-cacity is over 1GB.
+Being aware that the read and write latencies of NVM are asymmetric, we consider the read latency at first. In NVsim, we divide the entire cache read latency into the following components:
+\begin{itemize}
+\item (a)	H-tree input delay
+\item (b)	Decoder + word-line delay
+\item (c)	Bit-line delay + Sense Amplifier delay
+\item (d)	Comparator Delay (for tag part only)
+\item (e)	H-tree output delay
+\end{itemize}
+H-tree latency (a) and (f) are mainly determined by the RC delay of global wires, which is positive proportional to the area of memory macro.
+Sensing delay (d) is related to the read noise margin of memory cell that is affected by off/on resistance ratio.
+Figure~\ref{fig:memory-latency} (a) illustrates the read latency of different memories.
+We can see that sensing delay dominates the read latency of NVM at small capacity so that PCRAM (with the largest resistance window) is faster than RRAM and MRAM (with the smallest resistance window).
+While H-tree delay unveils at large capacity so that RRAM (with the smallest cell size) becomes faster than PRAM and MRAM (with the biggest cell size).
+The read latency of SRAM bank will increase rapidly after 128MB due to large area.
+
+Write latency of NVM are almost dominated by the write pulse width.
+In this work we assume 10ns, 20ns, 100ns for MRAM, RRAM and PCRAM.
+While write latency of SRAM and eDRAM is a function of capacity, as similar to the read latency.
+The results in ~\ref{fig:memory-latency} (b) indicates that NVM is suitable for memory with large capacity.
+
+Figure~\ref{fig:memory-energy} has demonstrated the dynamic energy of different memory technologies when 20\% and 50\% write access are assumed.
+eDRAM will be better than SRAM in terms of dynamic energy after 16MB and this is verified by IBM Power7 L3 cache.
+The cross-point between NVM and SRAM/eDRAM is postponed for 50\% write than 20\% write.
+
+Based on these observations, we would like to design a bandwidth-aware hybrid memory hierarchy, which always provides the high memory bandwidth with the given capacity. For example,
+\begin{itemize}
+\item eDRAM has better latency and enegry than SRAM when capacity is larger than 16MB.
+\item MRAM is more competitive to SRAM and eDRAM when capacity is larger than 128MB.
+\item PCRAM has serious endurance issue and is targeted as main memory replacement.
+\item RRAM might fit into cache hierarchy as last level cache replacement when there multiple levels of cache for applications with very few writes.
+\end{itemize}
 
@@
 
+We use Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator in our
+experiments. It is configured to model an eight-core CMP. Each core is in-order, and
+is similar to UltraSPARC III architecture. The frequency of each core is set to
+be 1GHz. Table~\ref{table:cmp-config} lists the detailed parameters of the
+baseline. Our baseline contains 8 private L1 instruction and data caches
+respectively. Since we only evaluate shared cache hierarchies, each of L1 cache
+is fixed to be SRAM-based, and of 64KB capacity. As regard to lower level
+caches, we evaluate two cases, pure SRAM-based and hybrid caches with various
+memory technologies, including SRAM, MRAM, RRAM, and eDRAM. By evaluating both
+cases, we would like to find out optimal cache design that leads to the peak
+performance, i.e., the number of cache levels that is reqired, memory technology
+used for each level, and capacity of each level.
+
+\begin{table}[t]
+  \centering
+  \caption{Baseline CMP configuration.}
+  \vspace{0.1in}
+  \begin{tabular}{l|c}
+    \hline
+    \multicolumn{2}{l}{\textbf{Core}}\\
+    \hline
+    No. of cores & 8\\
+    \hline
+    Frequency & 1GHz\\
+    \hline
+    Core architecture & in-order, 14-stage pipline, 2 ALUs, 2 FPUs\\
+    \hline
+    \multicolumn{2}{l}{\textbf{Memory}}\\
+    \hline
+    \multirow{2}{*}{Private caches} & L1 I-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
+                                    & L1 D-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
+    \hline
+    \multirow{2}{*}{Shared caches} & Case 1: Pure SRAM, 64B line, 8-way, write-back \\
+                                   & Case 2: Hybrid~(SRAM, MRAM, RRAM, eDRAM), 64B line, 8-way, write-back\\
+    \hline
+    Main memory & 4GB\\
+    \hline
+  \end{tabular}
+  \label{table:cmp-config}
+\end{table}
+
+The benchmarks are selected from PARSEC benchmark suite~\cite{PARSEC:2008} with
+multithreaded programs, which focus on emerging workloads that are designed to
+be representative of next-generation shared-memory programs for CMPs. Since the
+performance of different memory technologies are closely related to read and
+write intensities, we selected some workloads that vary in the average numbers
+of L2 cache read per thousand instructions~(RPKI) and write per thousand
+instructions~(WPKI), which are listed in
+Table~\ref{table:benchmarks}.\vspace{0.3in}
+
+\begin{table}[t]
+  \centering
+  \caption{Characteristics of selected benchmarks.}
+  \vspace{0.1in}
+  \begin{tabular}{l|c|c||l|c|c}
+    \hline
+    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI} &
+    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI}\\
+    \hline
+    blackscholes & 22.8 & 61.7 & ferret & 5.7 & 173.4\\
+    \hline
+    bodytrack & 5.4 & 139.5 & fluidanimate & 2.6 & 70.1\\
+    \hline
+    canneal & 5.4 & 29.3 & streamcluster & 17.3 & 16.9\\
+    \hline
+    facesim & 6.0 & 102.4 & swaptions & 2.6 & 121.4\\
+    \hline
+  \end{tabular}
+  \label{table:benchmarks}
+\vspace{0.5in}
+\end{table}
+
 \subsection{Results}
 
+We evaluate various possible configurations with shared caches by simulation,
+i.e., with possible numbers of levels, memory technologies, and cache capacities
+for each cache level. We consider SRAM and eDRAM as the possible memory
+technologies to implement L2 and L3 caches. Both MRAM and RRAM have higher write
+latency than SRAM. Furthermore, the endurance of RRAM is too low to be used as
+lower level caches. Consequently, these two memory technologies are only
+considered to be used as the last level cache.
+
+In the first set of experiments, we evaluate the system performance with
+two-level shared caches. Figure~\ref{fig:thput-l2l3} shows the system throughput
+with all the benchmarks. It is illustrated that implementing hybrid cache with
+eDRAM as the L3 cache helps with performance among most of the benchmarks. Such
+performance improvement is more than 10\% in Figure~\ref{fig:thput-l2l3}(a). It
+is indicated by our memory technology exploration that eDRAM shows more latency
+benefits with larger capacities. As a result, a larger eDRAM-based L3 cache
+leads to more performance improvement to the pure SRAM implementation, as
+illustrated by Figure~\ref{fig:thput-l2l3}(b). However, hybrid cache does not
+always improve the performance, as shown in Figure~\ref{fig:benchmark-l2l3}.
+With the benchmark \emph{canneal}, hybrid cache configurations outperform the
+pure SRAM implementation among overall range of various capacities. With the
+benchmark \emph{ferret}, however, pure SRAM implementation results in higher
+performance than hybrid cache implementation with the same capacity
+configuration.
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=3in]{figures/thput-l2l3-1}\hspace{0.2in}
+\includegraphics[width=3in]{figures/thput-l2l3-2}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+\caption{Performance comparison with two-level caches among various benchmarks.
+  (a) The L3 cache capacity is 16MB. (b) The L3 cache capacity is 32MB.}
+\label{fig:thput-l2l3}
+\end{figure*}
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=3in]{figures/canneal-l2l3}\hspace{0.2in}
+\includegraphics[width=3in]{figures/ferret-l2l3}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+\caption{Performance comparison with two-level caches among various cache
+  capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
+\label{fig:benchmark-l2l3}
+\end{figure*}
+
+In the second set of experiments, we evaluate the system performance with
+three-level shared caches. When evaluating hybrid cache configurations, we
+consider implementing the last level cache by MRAM and RRAM memory technologies,
+since the data transaction intensity is relatively low in the last level cache.
+Figure~\ref{fig:thput-l2l3l4} shows the results.
+Figure~\ref{fig:thput-l2l3l4}(a) illustrated that the performance more than 15\%
+higher on average with a last level cache implemented by MRAM than pure SRAM
+implementations. More performance improvement is obtained by increasing the last
+level cache capacity, as shown in Figure~\ref{fig:thput-l2l3l4}(b).
+Figure~\ref{fig:benchmark-l2l3l4} compares performance with pure SRAM and hybrid
+cache implementations with different capacities. In this case, hybrid cache
+implementation leads to higher performance in both benchmarks. The results of
+the two sets of experiments are illustrated together in
+Figure~\ref{fig:benchmark-overall}. An interesting observation is that hybrid
+cache implementation shows high performance improvement with one of the
+applications \emph{canneal}, whereas as little improvement with the other
+\emph{ferret}. The reason is that write latency of both MRAM and RRAM is higher
+than SRAM when capacity is less than 1GB. Performance of applications with large
+number of writes, such as \emph{ferret} is therefore not benefit from hybrid
+cache design.
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=3in]{figures/thput-l2l3l4-1}\hspace{0.2in}
+\includegraphics[width=3in]{figures/thput-l2l3l4-2}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+
+\caption{Performance comparison with three-level caches among various
+  benchmarks. (a) The L3 cache capacity is 16MB, and the L4 cache capacity is
+  512MB. (b) The L3 cache capacity is 32MB, and the L4 cache capacity is 1GB.}
+\label{fig:thput-l2l3l4}
+\end{figure*}
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=3in]{figures/canneal-l2l3l4}\hspace{0.2in}
+\includegraphics[width=3in]{figures/ferret-l2l3l4}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+\caption{Performance comparison with three-level caches among various cache
+  capacity. (a) The canneal benchmark. (b) The ferret benchmark..}
+\label{fig:benchmark-l2l3l4}
+\end{figure*}
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=3in]{figures/canneal}\hspace{0.2in}
+\includegraphics[width=3in]{figures/ferret}\\
+\hspace{0.02in}
+\makebox[3in][l]{\bf (a)}
+\makebox[3in][l]{\bf (b)}
+
+\caption{Performance comparison with both two-level and three-level caches among
+  various cache capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
+\label{fig:benchmark-overall}
+\end{figure*}
+
+Finally, we evaluate all the possible cache hierarchy configurations with
+exhaustive simulations. The optimal cache configurations of each benchmark is
+shown in Table~\ref{table:results}. Since the characteristics vary among the
+benchmarks, the optimal cache configurations are different among each of
+benchmarks. Applications with high write intensities, such as \emph{bodytrack},
+\emph{ferret}, and \emph{swaptions} tend to favor SRAM-based caches. Other
+applications benefit more from hybrid cache implementation in terms of
+performance.
+
+\begin{table}[htbp!]
+  \centering
+  \caption{Optimal cache configurations for each benchmark.}
+  \vspace{0.1in}
+  \begin{tabular}{|l|c|c|c|}
+    \hline
+    \textbf{Benchmarks} & \textbf{L2 Configuration} & \textbf{L3 Configuration} & \textbf{L4 Configuration}\\
+    \hline
+    blackscholes & SRAM, 8MB & eDRAM, 32MB & RRAM, 1GB\\
+    \hline
+    bodytrack & SRAM, 8MB & eDRAM, 128MB & -\\
+    \hline
+    canneal & SRAM, 8MB & eDRAM, 512MB & MRAM, 1GB\\
+    \hline
+    facesim & SRAM, 8MB & eDRAM, 32MB & MRAM, 1GB\\
+    \hline
+    ferret & SRAM, 8MB & eDRAM, 32MB & -\\
+    \hline
+    fluidanimate & SRAM, 8MB & eDRAM, 1GB & -\\
+    \hline
+    streamcluster & SRAM, 16MB & eDRAM, 128MB & MRAM, 512MB\\
+    \hline
+    swaptions & SRAM, 8MB & SRAM, 32MB & SRAM, 1GB\\
+    \hline
+  \end{tabular}
+  \label{table:results}
+\end{table}
+
 \section{Conclusion and Future Work}
+In this project, we explore performance and energy characteristics of various
+emerging memory technolgies. Based on our exploration, we evaluate the shared
+cache hierarchy design of CMPs optimized for performance by filling the off-chip
+memory bandwidth gap. According to our evaluation, SRAM-based caches leads to
+reasonable performance with write-intensive applications. Hybrid cache design
+help with performance with other types of benchmarks.
 
-In addition, we would like to put an upper bound to the total dynamic energy
-consumption of the memory hierarchy. Figure~\ref{fig:memory-energy} shows the
-estimation of dynamic power consumptions of different memory technologies with
-the provided bandwidths. In figure~\ref{fig:memory-energy}(c), we estimate the total
-dynamic energy with 10\% write activities.
+Furthermore, a variety of continuing studies is remained to be explored related
+to bandwidth-aware memory hierarchy design. First of all, we only examine cache
+hierarchy design currently. It is necessary to extend our exploration to the
+overall memory hierarchy design. On the other hand, we do not consider energy
+efficiency in this project. We would like to put some power constraints to the
+memory hierarchy, and examine the energy efficiency of various memory hierarchy
+configurations.
 
